### 2013-03-21 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding mandc paper to dissertation

To demonstrate this, consider the example adapted from Mervin's work
with Wagner and others in the same area \citep{mervin_variance_2012}
and presented in Figure~\ref{fig:msod_example}.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/mc_background/msod_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Overlapping domain example illustrating how domain
      overlap can reduce communication costs.}
    \textit{All particles start in the blue region of interest. The
      dashed line represents 0.5 domain overlap between domains.}}
  \label{fig:msod_example}
\end{figure}
In this example, 3 particle histories are presented emanating from the
blue region of interest. Starting with particle A, if no domain
overlap is used then the only the blue domain exists on the starting
processor. Particle A is then transported through 3 other domains
before the history ends, therefore requiring three communications to
occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
shown by the dashed line, then the starting process owns enough of the
domain such that no communications must occur in order to complete the
particle A transport process. Using 0.5 domain overlap also easily
eliminates cases such as the represented by the path of particle C. In
this case, particle C is scattering between two adjacent domains,
incurring a large latency cost for a single particle. Finally, with
particle B we observe that 0.5 domain overlap will still not eliminate
all communications. However, if 1 domain overlap were used, the entire
geometry shown in Figure~\ref{fig:msod_example} would be contained on
the source processor and therefore transport of all 3 particles
without communication would occur.

@@

%%---------------------------------------------------------------------------%%
\section{Domain-to-Domain Communication}
\label{sec:mc_comm_costs}
With Wagner's work, we observed that domain overlap implemented with a
domain decomposed Monte Carlo setting reduces the amount of nearest
neighbor communication that must occur during transport. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations without domain
overlap \citep{siegel_analysis_2012}. They arrive at a few empirical
relationships and generate a numerical experiment that provides
insight into the feasibility of large scale domain decomposed Monte
Carlo as well as the relationship of communication to system
parameters. In their work, they define a simple 3-dimensional system
subdivded into Cartesian block domains with each processor containing
a single domain. Given this geometry, a particle can either be
absorbed in its current domain or be transported to one of the 6
adjacent domains owned by another process.

To quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\lambda$, as:
\begin{equation}
  \lambda =
  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
    of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
  \label{eq:leakage_fraction}
\end{equation}
For their studies, they assumed that the value of $\lambda$ was
dependent on the total cross section of the system and the same for
all processes for simplicity and load balancing. Using this
relationship and their simple geometry, they were able to relate the
amount of compute time required to transport all source particles to
the number of source particles, the leakage fraction, the latency and
bandwidth of communication, the ratio of local to global geometry
size, and the total cross section of the system.

What they found was that there was indeed an optimum ratio of local
domain to global domain size that results in a minimum total compute
time. They also found that this ratio was only dependent on the
latency and bandwidth constraints of the parallel machine. We expect
this because nearest neighbor communication patterns scale with the
volume and surface sizes of the local domains. The larger the volume,
the more compute time is spent within the local volume and therefore
at some point you are limited by the latency of the system, or the
time it takes to receive the data to operate on. However, at a certain
point, when the surface of those volumes becomes large enough, the
amount of data that must be communicated grows large enough that it
may overcome the bandwidth limitations of the system, resulting in a
decrease in performance. In their numerical experiments, they used the
above geometry and leakage rates based on that computed for a reactor
system for a single material problem with a total cross section
equivalent to that of homogenized light water reactor. Given the
optimal local domain size, they observed for load balanced situations
that nearest neighbor communication costs during transport are a small
fraction of total compute time, leading to the conclusion that domain
decomposition is feasible for large-scale Monte Carlo calculations.

\subsection{Load Balancing Concerns}
\label{subsec:mc_load_balancing}
Although domain decomposition was shown to be efficient in a perfectly
load balanced situation in Siegel's work \citep{siegel_analysis_2012},
careful consideration must be made for situations where this is not
the case. Given the stochastic nature of the problem and lack of a
globally homogeneous domain, parallel Monte Carlo simulations are
inherently load imbalanced. Procassini and others worked to alleviate
some of the load imbalances that are generated by both particle and
spatial parallelism and are therefore applicable to the MSOD algorithm
\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
in which the number of times a particular domain was replicated was
dependent on the amount of work in that domain (i.e. domains with a
high particle flux and therefore more particle histories to compute
require more work). In this variation, domains that require more work
will be replicated more frequently at reduced particle counts in each
replication. Furthermore, Procassini and colleagues noted that as the
simulation progressed and particles were transported throughout the
domain, the amount of replication for each domain would vary as
particle histories began to diffuse, causing some regions to have
higher work loads and some to have smaller work loads than the initial
conditions.

Consider the example in Figure~\ref{fig:procassini_example} adapted
from Procassini's work.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/mc_background/procassini_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Example illustrating how domain decomposition can
      create load balance issues in Monte Carlo.}  \textit{A domain is
      decomposed into 4 zones on 8 processors with a point source in
      the lower left zone. As the particles diffuse from the source in
      the random walk sequence as shown in the top row, their tracks
      populate the entire domain. As given in the bottom row, as the
      global percentage of particles increases in a zone, that zone's
      replication count is increased.}}
  \label{fig:procassini_example}
\end{figure}
In this example, a geometry is decomposed into 4 domains on 8
processors with a point source in the bottom left domain. To begin,
because the point source is concentrated in one domain, that domain is
replicated 5 times such the amount of work it has to do per processor
is roughly balanced with the others. As the particles begin to diffuse
away from the point source, the amount of replication is adjusted to
maintain load balance. Near the end of the simulation, the diffusion
of particles is enough that all domains have equal replication.  By
doing this, load balance is improved as each domain has approximately
equal work although each domain may represent a different spatial
location and have a differing number of histories to
compute. Compared to Wagner's work where the fission source was
distributed relatively evenly throughout the domain, fixed source
problems (and especially those that have a point-like source) like
those presented in Procassini's work will be more prone to changing
load balance requirements.

\subsection{Reproducible Domain Decomposed Results}
\label{subsec:reproducible_mc}
The 2006 work of Brunner is notable in that the Monte Carlo codes used
to implement and test the algorithms adhered to a strict policy of
generating identical results independent of domain decomposition or
domain replication as derived from the work of Gentile and colleagues
\citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
given for obtaining results reproducible to machine precision for an
arbitrary number of processors and domains. Differences can arise from
using a different random number sequence in each domain and performing
a sequence of floating point operations on identical data in a
different order, leading to variations in round-off error and
ultimately a non-identical answer. They use a simple example,
recreated below in Figure~\ref{fig:gentile_example}, that illustrates
these issues.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/mc_background/gentile_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Gentile's example illustrating how domain
      decomposition can create reproducibility issues in Monte Carlo.}
    \textit{Both particles A and B start in zone 1 on processor
      1. Particle A moves to zone 2 on processor 2 and scatters back
      to zone 1 while B scatters in zone 1 and remains there. A1 and
      A2 denote the track of particle A that is in zone 1 while B1 and
      B2 denote the track of particle B that is in zone 1.}}
  \label{fig:gentile_example}
\end{figure}
In this example, the domain is decomposed on two processors with each
processor owning one of the two zones. Starting with particle A, it is
born in zone 1 and is transported to zone 2 where a scattering event
occurs. Concerning the first reproducibility issue, if the same
sequence of random numbers is not used to compute the trajectory from
the new scattering event, we cannot expect to achieve the same result
if the domain were not decomposed. If the numbers are different, the
scattering event in zone 2 may keep the particle there or even eject
it from the domain if the sequence were different. The second issue is
demonstrated by adding another particle B that remains in the
domain. In this case, an efficient algorithm will transport particle A
on processor 1 until it leaves zone 1 and then transport particle
B. Particle A will not renter the domain until it has been
communicated to processor 2, processor 2 performs the transport, and
it is communicated back to processor 1. If we are doing a track-length
tally in zone 1, then we sum the tracks lengths observed in that
zone. In the single processor, single zone case particle A would be
transported in its entirety and then particle B transported. This
would result in a tally sum with the following order of operations:
$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
of floating point operations, we cannot expect these to have an
identical result to machine precision as round-off approximations will
differ resulting in non-commutative addition.

>>>>

To demonstrate this, consider the example adapted from Mervin's work
with Wagner and others in the same area \citep{mervin_variance_2012}
and presented in Figure~\ref{fig:msod_example}.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/msod_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Overlapping domain example illustrating how domain
      overlap can reduce communication costs.}
    \textit{All particles start in the blue region of interest. The
      dashed line represents 0.5 domain overlap between domains.}}
  \label{fig:msod_example}
\end{figure}
In this example, 3 particle histories are presented emanating from the
blue region of interest. Starting with particle A, if no domain
overlap is used then the only the blue domain exists on the starting
processor. Particle A is then transported through 3 other domains
before the history ends, therefore requiring three communications to
occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
shown by the dashed line, then the starting process owns enough of the
domain such that no communications must occur in order to complete the
particle A transport process. Using 0.5 domain overlap also easily
eliminates cases such as the represented by the path of particle C. In
this case, particle C is scattering between two adjacent domains,
incurring a large latency cost for a single particle. Finally, with
particle B we observe that 0.5 domain overlap will still not eliminate
all communications. However, if 1 domain overlap were used, the entire
geometry shown in Figure~\ref{fig:msod_example} would be contained on
the source processor and therefore transport of all 3 particles
without communication would occur.

@@

%%---------------------------------------------------------------------------%%
\section{Domain-to-Domain Communication}
\label{sec:mc_comm_costs}
With Wagner's work, we observed that domain overlap implemented with a
domain decomposed Monte Carlo setting reduces the amount of nearest
neighbor communication that must occur during transport.

To date, parallel Neumann-Ulam methods have been limited to full
domain replication with parallelism exploited through individual
histories \cite{alexandrov_efficient_1998}. In reactor physics Monte
Carlo applications, however, domain decomposition has been identified
as a key principle in moving forward in high performance computing to
enable higher fidelity simulations
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because these nearest-neighbor
communication sequences, specifically, have poor algorithmic strong
scaling \cite{gropp_high-performance_2001}.

The purpose of this study is to provide a simple, analytic theory
based on the properties of the linear system that will allow for
estimates of the domain decomposed behavior of the adjoint
Neumann-Ulam method. When solving problems where the linear operator
is symmetric, a host of analytic theories exist based on the
eigenvalue spectrum of the operator that characterize their behavior
in the context of deterministic linear solvers. Using past work, these
theories are adapted to the domain decomposed adjoint Neumann-Ulam
method using the one-speed, two-dimensional neutron diffusion
equation. In this paper we describe the adjoint Neumann-Ulam Monte
Carlo method followed by a presentation of the model problem. Using
the linear system generated by the discretization of the model
problem, we use a spectral analysis to generate analytic relations for
the eigenvalues of the operator based on system parameters. Using the
eigenvalue spectra, we then build relationships to characterize the
transport of stochastic histories in a decomposed domain and the
fraction of histories that leak from a domain and will therefore have
to be communicated. Finally, we compare these analytic results to
numerical experiments conducted with the model problem and draw
conclusions looking towards future work.

\subsection{Model Problem}
\label{subsec:model_problem}
For our numerical experiments, we choose the one-speed,
two-dimensional neutron diffusion equation as a model problem
\cite{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
\cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.25}{\input{chapters/parallel_mc/stencil.pdftex_t}}
  \end{center}
  \caption{\textbf{Nine-point Laplacian stencil.}}
  \label{fig:stencil}
\end{figure}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.

To close the system, a set of boundary conditions is required. In the
case of a non-reentrant current condition applied to all global
boundaries of the domain, we choose the formulation of Duderstadt by
assuming the flux is zero at some ghost point beyond the
grid. Consider for example the equations on the $i=0$ boundary of the
domain:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
  \label{eq:x_min_bnd}
\end{multline}
Here we note some terms where $i=-1$ and therefore are representative
of grid points beyond the boundary of the domain. We set the flux at
these points to be zero, giving a valid set of equations for the $i=0$
boundary:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
  \label{eq:x_min_bnd_2}
\end{multline}
We repeat this procedure for the other boundaries of the domain. For
reflecting boundary conditions, the net current across a boundary is
zero.

\subsection{Spectral Analysis}
\label{subsec:spectral_analysis}
The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the eigenvalues of the iteration matrix. We
will compute these eigenvalues by assuming eigenfunctions of the form
\cite{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these eigenfunctions should
give the eigenvalues of the matrix as they exist on the unit circle in
the complex plane.

\subsection{Iteration Matrix Spectrum}
\label{subsec:iteration_spectrum}
For the model problem, we first compute the eigenvalues for the
diffusion operator $\ve{D}$ by applying the operator to the
eigenfunctions and noting that $x=ih$ and $y=jh$:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
  \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
preconditioning is sufficient to reduce the spectral radius of the
iteration matrix below unity and therefore ensure convergence of the
Neumann series. The preconditioner in this case is then $\ve{M} =
diag(\ve{D})$ such that we are solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
operator with each row scaled by the diagonal component. As we have
defined a homogeneous domain, the scaling factor, $\alpha$, is the
same for all rows in the operator and defined as the $\phi_{i,j}$
coefficient from Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

The spectral radius of the iteration matrix is obtained by seeking its
largest eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned iteration matrix is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{equation}
  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
    \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}]\:.
  \label{eq:iteration_stencil}
\end{equation}
Inserting the eigenfunctions defined by
Eq~(\ref{eq:eigenfunction_form}) we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. To find these maxium eigenvalue,
Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
values of $\Sigma_a$.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{chapters/parallel_mc/diffusion_spectrum.png}
  \end{center}
  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
  \label{fig:diffusion_spectrum}
\end{figure}.
 We find that the maximum eigenvalue exists when $p=q=0$,
giving the following for the spectral radius of the Jacobi
preconditioned iteration matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}

\subsection{Neumann Series Convergence}
\label{subsec:neumann_convergence}
The adjoint Monte Carlo method is effectively an approximation to a
stationary method. Stationary methods for linear systems arise from
splitting the operator in Eq~(\ref{eq:linear_problem}) and iterating:
\begin{equation}
  \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
  index}. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution
error at the $k^{th}$ iterate, the error after $k$ iterations is then:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
By assuming $\ve{H}$ is diagonalizable \cite{leveque_finite_2007}, we
then have:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}

In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
applications of the iteration matrix, are approximated by a random
walk of average length $k$ to yield the summation in
Eq~(\ref{eq:adjoint_neumann_solution})
\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
length, or the number of transitions before the termination of a
history (either by the weight cutoff, absorption, or exiting the
global domain) is therefore approximately the number of stationary
iterations required to converge to the specified tolerance. In the
case of the adjoint Neumann-Ulam method, no such tolerance exists,
however, we have specified a weight cutoff, $W_c$, that determines
when low-weight histories will be prematurely terminated as their
contributions are deemed minute. After $k$ iterations, a stationary
method is terminated as the error has reached some fraction,
$\epsilon$, of the initial error:
\begin{equation}
  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm4}
\end{equation}
Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
Neumann-Ulam method, if we take this fraction to be the weight cutoff,
a measure of how accurately the contributions of a particular history
to the solution are tallied, we then have the following relationship
for $k$:
\begin{equation}
  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
  \label{eq:analytic_k}
\end{equation}
This then gives us a means to estimate the length of the random walks
that will be generated from a particular linear operator based on the
eigenvalues of its iteration matrix (independent of the linear
operator splitting chosen) and based on the weight cutoff parameter
used in the Neumann-Ulam method.

\subsubsection{Domain Leakage Approximations}
\label{subsubsec:domain_leak_approx}
In a domain decomposed situation, not all histories will remain within
the domain they started in and must instead be communicated. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations for full nuclear
reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\Lambda$, as:
\begin{equation}
  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
          \label{eq:leakage_fraction}
\end{equation}
For their studies, Siegel and colleagues assumed that the value of
$\Lambda$ was dependent on the total cross section of the system via
the Wigner rational approximation. Outlined more thoroughly by Hwang's
chapter in \cite{azmy_nuclear_2010}, we will use both the Wigner
rational approximation and the mean chord approximation as a means to
estimate the leakage fraction.

In the case of domain decomposed linear operator equations, we can use
diffusion theory to estimate the optical thickness of a domain in the
decomposition and the corresponding leakage fraction in terms of
properties of the linear operator and the discretization. To begin we
must first calculate the mean distance a Monte Carlo history will move
in the grid by computing the mean squared distance of its movement
along the chord of length $l$ defined across the domain. After a
single transition a history will have moved a mean squared distance
of:
\begin{equation}
  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
  \label{eq:step_1_length}
\end{equation}
where $h$ is the size of the discrete grid elements along the chord
and $n_s$ is the number of grid elements a history will move on
average every transition. For our diffusion model problem, $n_s$ would
equate to the expected number of states in the $i$ (or $j$ as the problem is
symmetric) direction that a history will move in a single
transition and is dependent on the stencil used for the
discretization. After $k$ transitions in the random walk, the history
will have moved a mean squared distance of:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
  \label{eq:step_k_length}
\end{equation}
If our chord is of length $l$ and there are $n_i$ grid elements (or
states to which a history may transition) along that chord, then $h =
l / n_i$ giving:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
  \label{eq:step_k_length_sub}
\end{equation}
From diffusion theory, we expect the average number of interactions
along the chord to be:
\begin{equation}
  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
  \label{eq:optical_thickness_1}
\end{equation}
where $d$ is the dimensionality of the problem and $\sqrt{\langle
  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
Carlo history in the domain. We can readily interpret $\tau$ to be the
\textit{effective optical thickness} of a domain of length
$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
  \label{eq:optical_thickness_2}
\end{equation}
which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
relation for the effective optical thickness:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s}
  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
  \label{eq:optical_thickness_3}
\end{equation}

For optically thin domains, we expect that most histories will be
communicated, while optically thick domains will leak the fraction of
histories that did not interact within. Using the optical thickness
defined in Eq~(\ref{eq:optical_thickness_3}), we can then complete the
leakage approximations by defining the bounds of $\tau \rightarrow 0,
\Lambda \rightarrow 1$ and $\tau \rightarrow \infty, \Lambda
\rightarrow \tau^{-1}$.  With these bounds we can then define the
leakage fraction out of a domain for the adjoint Neumann-Ulam method
using the Wigner rational approximation:
\begin{equation}
  \Lambda = \frac{1}{1+\tau}\:,
  \label{eq:wigner_domain_leakage}
\end{equation}
and using the mean-chord approximation:
\begin{equation}
  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
  \label{eq:mean_chord_domain_leakage}
\end{equation}
Here, the leakage fraction is explicitly bound to the eigenvalues of
the iteration matrix, the size of the domain, the content of the
discretization stencil, and the weight cutoff selected to terminate
low weight histories.

\subsection{Numerical Experiments}
\label{subsec:numerical_experiments}
To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest eigenvalue of the diffusion operator
using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative eigenvalue
solver.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=4in,clip]{chapters/parallel_mc/measured_spec_rad.png}
    \end{center}
    \caption{Measured and analytic preconditioned diffusion operator
      spectral radius as a function of the absorption cross section to
      scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
      $h=1.0$ were used. The red data was computed numerically by an
      eigensolver while the black dashed data was generated by
      Eq~(\ref{eq:iteration_radius}).}
    \label{fig:measured_spec_rad}
  \end{spacing}
\end{figure}

\subsubsection{Random Walk Length}
\label{subsubsec:walk_length}
With the spectral data in hand, we can go about setting up an
experiment to measure the length of the random walks generated by the
adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
100$ square grid with $h=0.1$ and the absorption cross varied from 0
to 100 while the scattering cross section was fixed at unity. Three
weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
used with 10,000 histories generated by a point source of strength 1
in the center of the domain. For each of the histories, the number of
transitions made was tallied to provide an effective value of $k$ for
each history. This value was then averaged over all histories to get a
measured value of $k$ for the particular operator. On the left,
Figure~\ref{fig:measured_length} presents these measurements as well
as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
the right, Figure~\ref{fig:measured_length} gives the relative error
between the predicted and observed results. We note good qualitative
agreement between the measured and analytic results. However, we
observe a larger relative error for both long and short random walks.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=6.5in,clip]{chapters/parallel_mc/measured_length.png}
    \end{center}
    \caption{Measured and analytic random walk length as a function of
      the iteration matrix spectral radius. The weight cutoff was
      varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. In the left
      plot, the red data was computed numerically by an adjoint
      Neumann-Ulam implementation while the black dashed data was
      generated by Eq~(\ref{eq:analytic_k}). In the right plot, the
      relative error between the predicted and measured results is
      presented for each weight cutoff.}
    \label{fig:measured_length}
  \end{spacing}
\end{figure}
